### 2010-12-08 Documentation/atropos/atropos.tex
update to bib & update to tex - intro + abstract 'done'
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@691 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{singlespace} 
\scriptsize
\flushleft
%\line(1, 0){250} \\
{\bf Atropos:  $n$-Tissue Segmentation and Cortical Parcellation}\\
Corresponding author: \\
Brian B. Avants\\
3600 Market Street, Suite 370\\
Philadelphia, PA  19104\\
avants@picsl.upenn.edu\\
\end{singlespace}

@@

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm which uses either
parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  It is capable of incorporating prior labelings,
prior probability maps, and/or Markov Random Field (MRF) modeling
assumptions as spatial priors and geodesically propagating
segmentation information across neuroanatomy to ensure a dense
labeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) Markov Random
Field Segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based Segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London
{\bf FIXME proper name}.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} Neurological and neurodevelopmental research
rely on anatomical and functional correspondence both within and
between subjects.  As image acquisition technology has advanced,
concomitant investment has been made towards image classification
techniques for the cerebrum beginning the appropriation of NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
techniques has ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

Given the local optimization susceptibility characteristic of EM algorithms, various initialization schemes have been proposed.  Common low-level initialization strategies include uniform
probability assignment \citep{Wells1996}, Otsu thresholding \citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More sophisticated low-level initialization schemes include that of \cite{Greenspan2006} in which a dense spatial distribution of Gaussians is used to capture the complex neuroanatomical layout with subsequent processing used to conjoin subsets of Gaussians belonging to the same tissue classes.  Recently, anatomically specific strategies for incorporating prior knowledge  into the segmentation solution have included the use of spatial prior probability maps  of the various structures of interest \citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior probability maps can also be used to provide an
initial segmentation.   Related technological developments encompasses
those driven by partial volume modeling for increased accuracy in
brain segmentation \citep{Ruan2000,Ballester2002,Leemput2003}.

Observations regarding a general trend towards a more integrative view
of neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available as SPM---a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early seminal work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community. A similar holistic
approach to brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\cite{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions.  {\bf FIXME references proper names SIENA stuff}
Researchers in aging have also developed tools that have focused on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration.  Recent comparative evaluation studies {\bf
  FIXME cite a number of these} have shown that methods such as FAST,
SPM Unified Segmentation, SIENA and others show different performance
characteristics under different evaluation criteria.  That is, no
single method performs best under every measurement, highlighting
the need for segmentation tools that are tunable for different
problems and research goals.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various algorithms have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
Advantages of surface-based approaches include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, as well as Klein's
Mindboggle \cite{Klein2003}, would usually follow an initial segmentation by a volumetric method such as Atropos.

{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage of the pipeline we propose is that we parcellate
cortex entirely in the image space, thus avoiding the difficulty of
transferring labels from the mesh space back to the image space---a
problem that is a confound of surface-based methods\citep{Klein2010}.}}

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements the $n$-tissue paradigm for
volumetric image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
extend the use of spatial priors, in this work, to allow sparse priors
that may then be extended by a geodesic distance-based label propagation
approach.  In short, Atropos seeks to provide a toolbox of
segmentation approaches that may be modified, tuned and refined by the user
for different problems.

@@

The major contribution of the Atropos algorithm is a flexible, memory
efficient implementation of core expectation maximization machinery
that is useful across many segmentation problems.  To highlight the
value of this open-source, free contribution, we performed a search of
software attributes on NITRC and found that as of {\textcolor{red}
INSERT DATE NO STAND-ALONE N-CLASS EM methods are currently listed}.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term.  Such
an approach is valuable when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.

@@

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.
{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
  allow a weighted combination defined by the ComputeLocalPosterior
  function in Atropos --- how do we justify / explain this?   clearly,
it's a balance of the global solution (no local prior) and the local
one with spatial prior ... }}

>>>>

\begin{singlespace} 
\scriptsize
\flushleft
%\line(1, 0){250} \\
{\bf Atropos:  $n$-Tissue Segmentation}\\
Corresponding author: \\
Brian B. Avants\\
3600 Market Street, Suite 370\\
Philadelphia, PA  19104\\
avants@picsl.upenn.edu\\
\end{singlespace}

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based algorithmic approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling
assumptions.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) Markov Random
Field Segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based Segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} As medical image acquisition technology has advanced,
concomitant investment has been made towards adapting classification
techniques for neuroanatomy.  Some of the earliest work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

EM algorithms are sensitive to local optima. Consequently, a good initialization
strategy is critical.  Common low-level initialization strategies include uniform
probability assignment \citep{Wells1996}, Otsu thresholding
\citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More
sophisticated low-level initialization schemes include that of
\cite{Greenspan2006} in which a dense spatial distribution of
Gaussians is used to capture the complex neuroanatomical layout with
subsequent processing used to conjoin subsets of Gaussians belonging
to the same tissue classes.  Recently, anatomically specific
strategies for incorporating prior knowledge into the segmentation
solution have included the use of spatial prior probability maps of
the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments encompasses those driven by partial
volume modeling for increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5---a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community. A similar holistic
approach to brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions. Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via the SIENA algorithm
\citep{Smith2007}.  Two recent comparative evaluation studies \citep{,Bresser2011} have shown that methods such as FAST,
SPM Unified Segmentation, SIENA and others show different performance
characteristics under different evaluation criteria.  That is, no
single method performs best under every measurement, highlighting
the need for segmentation tools that are tunable for different
problems and research goals.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various algorithms have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
Advantages of surface-based approaches include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, as well as Klein's
Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.

%{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage
%of the pipeline we propose is that we parcellate cortex entirely in
%the image space, thus avoiding the difficulty of transferring labels
%from the mesh space back to the image space---a problem that is a confound of surface-based methods\citep{Klein2010}.}}

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
extend the use of spatial priors, in this work, to allow sparse priors
that may then be extended by a geodesic distance-based label propagation
approach.  In short, Atropos seeks to provide a toolbox of
segmentation approaches that may be modified, tuned and refined by the user
for different problems.

@@

The major contribution of the Atropos algorithm is a flexible,
well-designed, memory efficient implementation of core expectation maximization machinery
that is useful across many segmentation problems.  Atropos options
include different maximization criterion, an interactive mode, tunable
priors and multivariate density estimators using both
non-parametric and parametric functions.  To highlight the
value of this open-source, free contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

@@

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.  The user may also
choose mixing proportions equal to $\frac{1}{K}$ via the command line
interface to the posterior formulation.
{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
  allow a weighted combination defined by the ComputeLocalPosterior
  function in Atropos --- how do we justify / explain this?   clearly,
it's a balance of the global solution (no local prior) and the local
one with spatial prior ... }}
